Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
Pass on 4

To address these problems, we start by investigating examples, for which we provide solutions to problem 1 and get insights into what is achievable.

\section{Examples}\label{sec:example}
\subsection{Gaussian Variance}\label{ssec:gaussian-variance}

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m_*; \m_n)
	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptote~\eqref{eq:asymptote} .
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).

@@

\subsection{Gaussian}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, let's move on to the ubiquitous Gaussian $\cN(m,\sigma^2)$.
Gaussians actually offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, nor self-concordant (see App.\ref{app:gaussian}).
Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.

>>>>

To address these problems, we start by investigating simple examples for which we can provide solutions to Problem 1, getting insights into what is achievable.

\section{Illustrating Examples}\label{sec:example}
\subsection{Gaussian with Unknown Variance}\label{ssec:gaussian-variance}

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m^*; \m_n)
	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.~\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
Note that this inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}.
Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth. 
This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).

@@

\subsection{Full Gaussian (Non-Trivial)}
\label{ssec:gaussian}
Now that we have solved the case of $\cN(0,\sigma^2)$, let us consider the full Gaussian $\cN(m,\sigma^2)$ which actually offers a highly non-trivial example for Problem 1.
Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
With sufficient statistic $T(x)=(x, x^2)$,
the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
Given these parameters, log-partition and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
}

These functions are neither smooth, nor strongly convex, nor self-concordant (see App.~\ref{app:gaussian}).
Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.
